---
author: "Eric Boxer"
date: "January 27, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r load data}
library(dagitty)
library(ggplot2)
library(ggpubr)
library(ggridges)
library(MatchIt)
library(tidyverse)
df <- read_csv('resources/arts_demo_ela_math.csv')
theme_eric <- function(base_size = 11,
                      base_family = 'URWHelvetica',
                      base_line_size = base_size / 170,
                      base_rect_size = base_size / 170){
  theme_bw(base_size = base_size, 
                base_family = base_family,
                base_line_size = base_line_size) %+replace%
    theme(
      plot.title = element_text(
        color = rgb(25, 43, 65, maxColorValue = 255), 
        face = "bold",
        hjust = 0),
      axis.title = element_text(
        color = rgb(105, 105, 105, maxColorValue = 255),
        size = rel(0.75)),
      axis.text = element_text(
        color = rgb(105, 105, 105, maxColorValue = 255),
        size = rel(0.5)),
      panel.grid.major = element_line(
        rgb(105, 105, 105, maxColorValue = 255),
        linetype = "dotted"),   
      panel.grid.minor = element_line(
        rgb(105, 105, 105, maxColorValue = 255),
        linetype = "dotted", 
        size = rel(4)),
      
      complete = TRUE
    )
}
theme_set(theme_eric())
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
c1 <- '#000044'
c4 <- c('#222222', '#555555', '#888888', '#BBBBBB')
c5 <- c('#222222', '#444444', '#666666', '#888888', '#AAAAAA')
```

```{r q names}
q_arts <- df %>% colnames() %>% .[grepl("^Q", .)]
q_ela <- df %>% colnames() %>% .[grepl("ela$", .)]
q_math <- df %>% colnames() %>% .[grepl("math$", .)]
q_demo <- df %>% colnames() %>% setdiff(union(q_arts, union(q_ela, q_math)))
disc_4 <- c("Dance", "Music", "Theater", "Visual Arts")
disc_5 <- c("Dance", "Film", "Music", "Theater", "Visual Arts")
```

```{r get perc_34, results = 'hide'}
#Get performance metrics of interest, % of passing students
perc_34_ela <- q_ela %>% .[grepl("^perc_34", .)]
perc_34_math <- q_math %>% .[grepl("^perc_34", .)]

#Get percent passing for all grades in 2018.
perc_34_all_2018 <- union(perc_34_ela, perc_34_math) %>%
  .[grepl("All Grades_2018", .)]
df %>% dplyr::select(perc_34_all_2018) %>% is.na() %>% colSums()
#Missing data for 370 schools
#Get percent passing for 4th graders in 2018
perc_34_4_2018 <- union(perc_34_ela, perc_34_math) %>% 
  .[grepl("4_2018", .)]
df %>% dplyr::select(perc_34_4_2018) %>% is.na() %>% colSums()

#Rename columns
temp <- c('perc_34_all_2018_ela', 'perc_34_all_2018_math')
colnames(df)[colnames(df) %in% perc_34_all_2018] <- temp
perc_34_all_2018 <- temp

perc_4_all_2018 <- df %>% colnames() %>% .[grepl('perc_4_', .)] %>% 
  .[grepl('_All Grades_2018', .)]
#Rename perc_4
temp <- c('perc_4_all_2018_ela', 'perc_4_all_2018_math')
colnames(df)[colnames(df) %in% perc_4_all_2018] <- temp
perc_4_all_2018 <- temp
```

```{r q3 missing, results = 'hide'}
#Q3 arts supervisor
q3 <- q_arts %>% .[grepl("^Q3_", .)]
df %>% dplyr::select(q3) %>% sapply(function(x) sum(is.na(x)))
```

### Arts Supervisors

Schools were asked whether their arts supervisor was employed full- or part-time, and if the supervisor was full-time whether they were solely working on arts programs or had other responsibilities.  
In small schools, or under-resourced ones, faculty may be expected to wear many hats. The simplest examples are physical education teachers leading gym classes and health classes. More extreme cases can have teachers with certification in, for example, English teaching students in math or science classes. I will not assume that an arts supervisor *must* be employed full-time to run an effective program, but I am curious as to what features predict their employment status.

```{r q3 bar}
superv_status <- factor(c('Full-time, solely for arts', 'Full-time, with other duties', 'Part-time', 'None'),
                          levels=c('Full-time, solely for arts', 'Full-time, with other duties', 'Part-time', 'None'))

df %>% dplyr::select(q3) %>%
  colSums() %>% as_data_frame() %>%
  ggplot(aes(x=superv_status, y=value)) +
  geom_bar(stat = 'identity') +
  ggtitle('Arts Supervisor Employment Status') +
  xlab('Employment Status') + ylab('Frequency') +
  theme(text = element_text(size=20),
        axis.text.x = element_text(size=8))
```

There are many full-time supervisors with duties other than the arts. I would be curious to discover the share of their responsibilities that are considered "other". It is possible that some administrators consider teaching an arts class or doing clerical work to be "other". Relative to the number of supervisors working full-time, there are few part-timers.  

```{r missing q3, results = 'hide'}
df$Q3 <- with(df, ifelse(Q3_1 == 1, '1',
                ifelse(Q3_2 == 1, '2',
                       ifelse(Q3_3 == 1, '3',
                              ifelse(Q3_4 == 1, '4', '0')))))
df$Q3 <- df$Q3 %>% factor(levels=seq(0,4))
df %>% filter(Q3 == '0') %>% count()
#Two schools did not respond to Q3
```

Only two schools did not respond to this question about their arts supervisor's status. In the previous part, I looked at a question that asked whether a school had a designated arts liaison. 72 schools responded that they did not, while 296 schools do not have any arts supervisor. I wonder whether the presence of an arts supervisor is more or less influential in student academic performance, compared to an arts liaison.

To begin exploring academic performance, I will use the percentage of students to perform at or above standards for English Language Arts (ELA) and math.

```{r math and ela scores}
df %>% dplyr::select(perc_34_all_2018, Q3) %>%
  filter(!is.na(perc_34_all_2018_ela)) %>% 
  filter(Q3 != 0) %>% 
  mutate(Q3 = factor(ifelse(Q3 == 1, 'Full-time, solely for arts',
                            ifelse(Q3 == 2, 'Full-time, with other duties',
                                   ifelse(Q3 == 3, 'Part-time', 'None'))))) %>% 
  ggplot(aes(x=perc_34_all_2018_ela, y=perc_34_all_2018_math)) +
  geom_point(alpha = .9) + facet_wrap(~Q3,
                            labeller = label_value) +
  ggtitle('Academic Performance, by Supervisor Status') +
  xlab('Students Passing, ELA (%)') +
  ylab('Students Passing, Math (%)') +
  theme(text = element_text(size=16),
        axis.text.x = element_text(size=8))
```

Coming into this analysis I had some lingering thoughts that arts programs might be more beneficial for ELA performance than for math, if they have any effect. In the plot above there is a positive relationship between ELA and math scores, regardless of art supervisor status.  

```{r supervisor status stats, results = 'hide'}
for (i in seq(4)) {
  lm.fit <- df %>% select(perc_34_all_2018, Q3) %>% 
    filter(Q3 == i) %>% 
    lm(perc_34_all_2018_ela~perc_34_all_2018_math, data=.)
  print(i)
  print(lm.fit %>% summary())
  print(lm.fit %>% confint())
}
```

```{r supervisor performance stats null}
lm.fit.null <- df %>% select(perc_34_all_2018) %>% 
  rename(Math_score = perc_34_all_2018_math,
         ELA_score = perc_34_all_2018_ela) %>% 
  lm(Math_score~., data=.)
lm.fit.null.summ <- lm.fit.null %>% summary()
lm.fit.null.summ
```

On inspection the relation between the two scores seems identical between supervisor statuses. In fact, fitting a linear regression to each yields coefficients close to the uncontrolled coefficient for math scores on ELA scores, 1.06415. $Math\_score = \beta_0 + \beta_1 * ELA\_score$  

```{r supervisor performance stats control}
lm.fit <- df %>% select(q3, perc_34_all_2018) %>% 
  rename(Math_score = perc_34_all_2018_math,
         ELA_score = perc_34_all_2018_ela) %>% 
  lm(Math_score~., data=.)
lm.fit.summ <- lm.fit %>% summary()
lm.fit.summ
```

We can control for supervisor status with dummy variables by altering the regression like so, $Math\_score = \beta_0 + \beta_1 * ELA\_score + \beta_2 * Q3\_1 + \beta_3 * Q3\_2 + \beta_4 * Q3\_3 + \beta_5 * Q3\_4$, where $Q3\_i$ corresponds to the $i^{th}$ member of the list `c('Full-time, solely for arts', 'Full-time, with other duties', 'Part-time', 'None')`.  

```{r supervisor performance diff, echo=TRUE}
delta.beta1 <- coefficients(lm.fit.null.summ)[2,1] - coefficients(lm.fit.summ)[6,1]
delta.beta1
se.delta.beta1 <- sqrt(coefficients(lm.fit.null.summ)[2,2]^2 + coefficients(lm.fit.summ)[6,2]^2)
se.delta.beta1
```

On inspection, the regression coefficients for `ELA_score` are close between model specifications A (no controls) and B (controlling for supervisor status). The difference in estimated coefficients $\beta_1$ is -0.0003. To find the variability of the difference in regression coefficients we use the formula $Var(A-B) = Var(A) + Var(B) - 2*Cov(A,B)$. Assuming the covariance in the two estimates is zero yields an 0.0201 as standard error of the difference.  
Now I feel confident in the claim that there is no difference in the relationsip between ELA and math scores from supervisor statuses.  

```{r mosaic}
library(ggmosaic)
df %>% select(Q2_1, Q3) %>% 
  mutate(Q2_1 = ifelse(Q2_1 == 1, 'No', 'Yes'),
         Q3 = factor(ifelse(Q3 == 1, 'Full-time, solely for arts',
                            ifelse(Q3 == 2, 'Full-time, with other duties',
                                   ifelse(Q3 == 3, 'Part-time', 'None'))))) %>% 
  mutate(Q3 = as_factor(Q3, levels = c('Full-time, solely for arts',
                    'Full-time, with other duties',
                    'Part-time',
                    'None'))) %>% 
  ggplot() + 
  geom_mosaic(aes(x=product(Q3, Q2_1), fill=Q3), na.rm=T) +
  scale_fill_manual(values=c5) +
  ggtitle('Does Having an Arts Liaison Affect Supervisor Status?') +
  xlab('School arts liaison?') +
  ylab('Arts Supervisor Status') +
  theme(text = element_text(size=16),
        axis.text.x = element_text(size=8))
```

The mosaic plot illustrates that the proportion of schools without an arts program supervisor is greater for schools without an arts liaison that for those with. This lends support to the idea that those schools are lacking the resources to fully staff their arts programs, as they have not filled two key positions.  

The next questions concern certifications that arts supervisors may have, either in an arts discipline or in administration.  

```{r supervisor cert both, results = 'hide'}
df %>% filter(Q4_1 == 1 & Q5_1 == 1) %>% count()
df %>% filter(Q4_1 == 1) %>% count()
df %>% filter(Q5_1 == 1) %>% count()
```

```{r supervisor cert}
df$supervisor_cert <- with(df, ifelse(Q4_1 == 1 & Q5_1 == 1, 'Both',
                ifelse(Q4_1 == 1 & Q5_1 == 0, 'Arts',
                       ifelse(Q4_1 == 0 & Q5_1 == 1, 'Administration', 'None'))))
df %>% dplyr::select(Q0_DBN, supervisor_cert) %>% 
  group_by(supervisor_cert) %>% 
  count() %>% 
  ggplot(aes(x=supervisor_cert, y=n)) +
  geom_bar(stat='identity') +
  ggtitle('Does the Arts Supervisor Have Any Certifications?') +
  xlab('Supervisor Certification') + ylab('Frequency') +
  theme(text = element_text(size=16),
        axis.text.x = element_text(size=8))
```

Few supervisors are certified in the arts, the majority are administrators. Could this have an impact on the efficacy of an arts program?  

```{r supervisor cert stats, results = 'hide'}
df <- df %>%
  mutate(supervisor_cert_both = ifelse(Q4_1 == 1 & Q5_1 == 1, 1, 0))
q_certs <- c('Q4_1', 'Q5_1', 'supervisor_cert_both')
df %>%
  dplyr::select(q_certs, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
df %>%
  dplyr::select(q_certs, perc_34_all_2018_math) %>% 
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
```

```{r supervisor cert stats ctrl, results = 'hide'}
df <- df %>%
  mutate(supervisor_cert_both = ifelse(Q4_1 == 1 & Q5_1 == 1, 1, 0))
q_certs <- c('Q4_1', 'Q5_1', 'supervisor_cert_both')
lm.ela <- df %>%
  dplyr::select(q_certs, total_enrollment_2017, perc_black_2017, perc_pov_2017, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.)
lm.ela %>% summary()
lm.ela %>% confint()
lm.math <- df %>%
  dplyr::select(q_certs, total_enrollment_2017, perc_black_2017, perc_pov_2017, perc_34_all_2018_math) %>% 
  lm(perc_34_all_2018_math~., data=.)
lm.math %>% summary()
lm.math %>% confint()
```

Linear regressions do not yield a statistically significant result for the effect of either or both supervisor certification on student academic performance.  
Controlling for school size and percentage of black and impoverished students, there is no statistically significant effect for either certification. A 95% confidence interval of the coefficient for supervisor certification in administration includes zero for ELA and math scores, so we cannot claim that there is a statistically significant effect for that certification on academic performance. The arts certification confidence interval is even wider.  

```{r supervisor cert stats narrow ctrl, results = 'hide'}
df <- df %>%
  mutate(supervisor_cert_both = ifelse(Q4_1 == 1 & Q5_1 == 1, 1, 0))
q_certs <- c('Q4_1', 'Q5_1', 'supervisor_cert_both')
lm.ela <- df %>%
  dplyr::select(q_certs, total_enrollment_2017, perc_black_2017, perc_pov_2017, perc_4_all_2018_ela) %>% 
  lm(perc_4_all_2018_ela~., data=.)
lm.ela %>% summary()
lm.ela %>% confint()
lm.math <- df %>%
  dplyr::select(q_certs, total_enrollment_2017, perc_black_2017, perc_pov_2017, perc_4_all_2018_math) %>% 
  lm(perc_4_all_2018_math~., data=.)
lm.math %>% summary()
lm.math %>% confint()
```

If we drill down further, looking at only the percentage of students to receive a 4 (the highest grade), then there is an almost statistically significant coefficient for the effect of arts certification on ELA scores, with a p-value of 0.0585. However, a 95% confidence interval does still include zero. Having been able to arrive at this more tenuous result I feel conflicted between a renewed belief that arts programs *do* have an effect on academic performance and suspicion as to how I *manufactured* it by narrowing my focus until I found an accceptable p-value.  

===

### Resources devoted to arts education

Schools were asked how many rooms they have dedicated to arts education. This may be useful for assessing the resources that are made available for arts education more broadly.

```{r rooms setup}
q8 <- df %>% colnames() %>% .[grepl('Q8_', .)]
df <- df %>%
  mutate(rm_dance = Q8_R1_C1 + Q8_R1_C2,
         rm_music = Q8_R2_C1 + Q8_R2_C2,
         rm_thtr = Q8_R3_C1 + Q8_R3_C2,
         rm_visart = Q8_R4_C1 + Q8_R4_C2,
         rm_media = Q8_R5_C1 + Q8_R5_C2)
rooms <- df %>% colnames %>% .[grepl('rm_', .)]
df %>% 
  dplyr::select(starts_with('rm')) %>% 
  gather(key = discipline,
         value = num) %>%
  ggplot(aes(x=num)) +
  geom_density() +
  facet_wrap(~discipline)
df %>% 
  dplyr::select(starts_with('rm')) %>% 
  gather(key = discipline,
         value = num) %>%
  ggplot(aes(x=num)) +
  geom_histogram(bins=30) +
  facet_wrap(~discipline)
```

All disciplines show the same trend, many schools have zero to two rooms for any discipline and a few schools have more than twenty rooms. Having twenty-some rooms strikes me as being unrealistic, but this was a representation of rooms designed solely for the arts and multi-purpose rooms. How about if we look at just rooms dedicated solely for the arts.

```{r rooms solely}
df %>% select(q8) %>%
  select(ends_with('C1')) %>% 
  gather(key = discipline,
         value = num) %>% 
  filter(!is.na(num)) %>% 
  ggplot(aes(x=num)) +
  geom_density() +
  facet_wrap(~discipline)
df %>% select(q8) %>%
  select(ends_with('C2')) %>% 
  gather(key = discipline,
         value = num) %>% 
  filter(!is.na(num)) %>% 
  ggplot(aes(x=num)) +
  geom_density() +
  facet_wrap(~discipline)
```

Looking at the number of rooms designed solely for the arts, the maximum number of rooms is a more reasonable 13. For a school with several hundred students and a dedicated arts program I can picture that. If we look at multi-purpose rooms, we can see that this is where the bulk of the rooms are from. Not many schools have dozens of multi-purpose rooms for arts disciplines, but again, in the context of a large school, it seems reasonable.  
I would be curious to discover the overlap for multi-purpose rooms among arts disciplines. In other words, is a school with twenty multi-purpose rooms reporting many of those rooms for each arts discipline? I do not see any way to completely correct for that possibility, but it would be interesting to control for school size to try to get to the bottom of it. If this does not work, then we could consider only the number of rooms designed solely for art, but this could penalize small schools unduly.  

```{r rooms stats}
#ELA
df %>% dplyr::select(rooms, total_enrollment_2017, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
#Math
df %>% dplyr::select(rooms, total_enrollment_2017, perc_34_all_2018_math) %>%
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
```

Conditioning on the effect of school size, through total enrollment, there is no statistically significant effect for the total number of rooms for arts education on student academic performance.  

```{r rooms stats}
rooms_ded <- df %>% select(q8) %>%
  select(ends_with('C1')) %>% colnames()
rooms_mul <- df %>% select(q8) %>%
  select(ends_with('C2')) %>% colnames()
#ELA dedicated
df %>% dplyr::select(rooms_ded, total_enrollment_2017, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
#Math dedicated
df %>% dplyr::select(rooms_ded, total_enrollment_2017, perc_34_all_2018_math) %>%
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
#ELA multi
df %>% dplyr::select(rooms_mul, total_enrollment_2017, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
#Math multi
df %>% dplyr::select(rooms_mul, total_enrollment_2017, perc_34_all_2018_math) %>%
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
```

Controlling for school size, through total student enrollment, we can see a statistically significant (p-value less than 0.001) positive coefficient for the effect of additional rooms designed and used solely for music, on ELA state test scores. If we consider math scores, rooms solely dedicated to music still have a positive effect, but at a lesser 0.0044 p-value. No other arts disciplines have a statistically significant effect on ELA scores, but dance and media arts rooms have statistically significant (p-value less than 0.05) negative effects on math scores.  
Considering multi-purpose rooms, no arts discipline has a statistically significant effect on academic performance.  

```{r rooms grades}
ela_2018 <- q_ela %>% .[grepl('perc_34_[3-8]_2018',.)]
math_2018 <- q_math %>% .[grepl('perc_34_[3-8]_2018',.)]
for (i in ela_2018) {
  temp <- df %>%
    dplyr::select(i, rooms_ded, total_enrollment_2017) %>% 
    .[complete.cases(.),]
  f.ela.1 <- paste(as.character(i), 'total_enrollment_2017', sep='~')
  f.ela.2 <-  paste(rooms_ded, collapse='+')
  f.ela <- as.formula(paste(f.ela.1, f.ela.2, sep='+'))
  temp <- temp %>% lm(f.ela, data=.)
  print(i)
  print(temp %>% summary())
  print(temp %>% confint.default())
}
for (i in math_2018) {
  temp <- df %>%
    dplyr::select(i, rooms_ded, total_enrollment_2017) %>% 
    .[complete.cases(.),]
  f.math.1 <- paste(as.character(i), 'total_enrollment_2017', sep='~')
  f.math.2 <-  paste(rooms_ded, collapse='+')
  f.math <- as.formula(paste(f.math.1, f.math.2, sep='+'))
  temp <- temp %>% lm(f.math, data=.)
  print(i)
  print(temp %>% summary())
  print(temp %>% confint.default())
}
```

If we break down academic performance by grade, then the results are not as straigtforward. For sixth-, seventh- and eighth-graders' ELA performance, media arts rooms have a more statistically significant negative coefficient than the positive coefficient for music rooms. All grades show a positive coefficient for music rooms, with a p-value no more than 0.01. There is a large jump in media arts between fifth- and sixth-grades, which is the demarcation between elementary and middle-schools. This leads me to believe that there is an underlying reason, beyond some increased susceptibility to media arts programs at the expense of academic performance. Perhaps not many elementary schools have media arts programs at all, or they are more commonly found in schools that are performing at an atypical level. Both of these can be checked. Note that the term Media Arts refers to film programs.  
Math scores show a similar trend to ELA scores, in that most grades have a statistically significant positive coefficient for music rooms. Among middle-schools, media arts rooms have a statistically significant negative coefficient. To me, the key difference is that there is not a statistically significant positive coefficient for music rooms on eighth-grade math scores.  

```{r media arts elem}
q_elem <- q_demo %>% .[grepl('grd_[1-5]_2017', .)]
q_mid <- q_demo %>% .[grepl('grd_[6-8]_2017', .)]
disc_5.1 <- c('Dance', 'Music', 'Theater', 'Visual Arts', 'Media Arts')
for (i in 1:5){
  temp <- df %>% select(q_elem[i], rooms_ded) %>% 
    filter((!!sym(q_elem[i])) > 0) %>% 
    select(-(!!sym(q_elem[i]))) %>% 
    colSums(na.rm = T) %>%
    as_data_frame() %>% 
    ggplot(aes(x=disc_5.1, y=value)) +
    geom_bar(stat='identity') +
    ggtitle(q_elem[i])
  print(temp)
}
for (i in 1:3){
  temp <- df %>% select(q_mid[i], rooms_ded) %>% 
    filter((!!sym(q_mid[i])) > 0) %>% 
    select(-(!!sym(q_mid[i]))) %>% 
    colSums(na.rm = T) %>%
    as_data_frame() %>% 
    ggplot(aes(x=disc_5.1, y=value)) +
    geom_bar(stat='identity') +
    ggtitle(q_mid[i])
  print(temp)
}
for (i in q_elem){
  temp <- df %>% select(i, rooms_ded) %>% 
    filter((!!sym(i)) > 0) %>% 
    select(-(!!sym(i))) %>% 
    gather(key = discipline,
           value = num) %>% 
    ggplot(aes(x=num)) +
    geom_density() +
    facet_wrap(~discipline) +
    ggtitle(i)
  print(temp)
}
for (i in q_mid){
  temp <- df %>% select(i, rooms_ded) %>% 
    filter((!!sym(i)) > 0) %>% 
    select(-(!!sym(i))) %>% 
    gather(key = discipline,
           value = num) %>% 
    ggplot(aes(x=num)) +
    geom_density() +
    facet_wrap(~discipline) +
    ggtitle(i)
  print(temp)
}
```

I do not see a significant difference in the distribution of art rooms between elementary- and middle-schools. Let's look at the performance of schools with media arts programs.  
We can match schools on having media arts rooms by controlling for number of students, number of students in each grade, and perhaps other demographic and academic features. Then we can look at effect on academic performance.  
```{r rooms media match}
library(MatchIt)
#Feature for having rooms dedicated to media arts
df <- df %>% mutate(rm_ded_media = min(1, Q8_R5_C1))
f.1 <- paste(paste(q_elem, collapse='+'), paste(q_mid, collapse='+'), sep='+')
f <- as.formula(paste('rm_ded_media~perc_34_all_2018_ela+total_enrollment_2017', f.1, sep='+'))
temp <- df %>%
  select(rm_ded_media, perc_34_all_2018_ela, total_enrollment_2017, q_elem, q_mid) %>% 
  .[complete.cases(.),]
m.out1 <- matchit(f, data=temp, method='nearest', distance='logit')
summary(m.out1)
plot(m.out1)

m.data1 <- match.data(m.out1, distance='pscore')
hist(m.data1$pscore, breaks=30)
summary(m.data1$pscore)

with(m.data1, t.test(perc_34_all_2018_ela ~ rm_ded_media))

lm.treat1 <- lm(perc_34_all_2018_ela~rm_ded_media, data=m.data1)
summary(lm.treat1)

lm.treat2 <- m.data1 %>% select(-weights) %>% 
  lm(perc_34_all_2018_ela~., .)
summary(lm.treat2)
```

Propensity score matching suggests that rooms dedicated to media arts are associated with schools with higher ELA scores. The result that led to this investigation was that rooms dedicated to media arts are associated with lower state test scores among middle-schoolers. In light of this analysis I would not draw any firm conclusions, but would be interested to find some instrumental variable to use in lieu of an experiment.  

## NYC Schools Arts Survey Data
The [2017-2018 Arts Survey Data](https://data.cityofnewyork.us/Education/2017-2018-Arts-Survey-Data/475h-cg5t) has data about arts teachers, budgets, partnerships with cultural organizations and parental involvement in NYC public schools.  

In an effort to gain greater context for this data, we can examine it in conjuction with publicly available [ELA and Math state test results](https://infohub.nyced.org/reports-and-policies/citywide-information-and-data/test-results) and [demographic data](https://data.cityofnewyork.us/Education/2013-2018-Demographic-Snapshot-School/s52a-8aq6).  

My goals were to understand the state of arts programs in NYC schools, what variables affect the resources of arts programs, and whether arts programs have an effect on the academic performance of students.  
