---
author: "Eric Boxer"
date: "January 23, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r load data}
library(dagitty)
library(ggplot2)
library(ggpubr)
library(ggridges)
library(MatchIt)
library(tidyverse)
df <- read_csv('resources/arts_demo_ela_math.csv')
theme_eric <- function(base_size = 11,
                      base_family = 'URWHelvetica',
                      base_line_size = base_size / 170,
                      base_rect_size = base_size / 170){
  theme_bw(base_size = base_size, 
                base_family = base_family,
                base_line_size = base_line_size) %+replace%
    theme(
      plot.title = element_text(
        color = rgb(25, 43, 65, maxColorValue = 255), 
        face = "bold",
        hjust = 0),
      axis.title = element_text(
        color = rgb(105, 105, 105, maxColorValue = 255),
        size = rel(0.75)),
      axis.text = element_text(
        color = rgb(105, 105, 105, maxColorValue = 255),
        size = rel(0.5)),
      panel.grid.major = element_line(
        rgb(105, 105, 105, maxColorValue = 255),
        linetype = "dotted"),   
      panel.grid.minor = element_line(
        rgb(105, 105, 105, maxColorValue = 255),
        linetype = "dotted", 
        size = rel(4)),
      
      complete = TRUE
    )
}
theme_set(theme_eric())
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
c1 <- '#000044'
c4 <- c('#222222', '#555555', '#888888', '#BBBBBB')
c5 <- c('#222222', '#444444', '#666666', '#888888', '#AAAAAA')
```

```{r q names}
q_arts <- df %>% colnames() %>% .[grepl("^Q", .)]
q_ela <- df %>% colnames() %>% .[grepl("ela$", .)]
q_math <- df %>% colnames() %>% .[grepl("math$", .)]
q_demo <- df %>% colnames() %>% setdiff(union(q_arts, union(q_ela, q_math)))
disc_4 <- c("Dance", "Music", "Theater", "Visual Arts")
disc_5 <- c("Dance", "Film", "Music", "Theater", "Visual Arts")
```

```{r get perc_34, results = 'hide'}
#Get performance metrics of interest, % of passing students
perc_34_ela <- q_ela %>% .[grepl("^perc_34", .)]
perc_34_math <- q_math %>% .[grepl("^perc_34", .)]

#Get percent passing for all grades in 2018.
perc_34_all_2018 <- union(perc_34_ela, perc_34_math) %>%
  .[grepl("All Grades_2018", .)]
df %>% dplyr::select(perc_34_all_2018) %>% is.na() %>% colSums()
#Missing data for 370 schools
#Get percent passing for 4th graders in 2018
perc_34_4_2018 <- union(perc_34_ela, perc_34_math) %>% 
  .[grepl("4_2018", .)]
df %>% dplyr::select(perc_34_4_2018) %>% is.na() %>% colSums()

#Rename columns
temp <- c('perc_34_all_2018_ela', 'perc_34_all_2018_math')
colnames(df)[colnames(df) %in% perc_34_all_2018] <- temp
perc_34_all_2018 <- temp

perc_4_all_2018 <- df %>% colnames() %>% .[grepl('perc_4_', .)] %>% 
  .[grepl('_All Grades_2018', .)]
#Rename perc_4
temp <- c('perc_4_all_2018_ela', 'perc_4_all_2018_math')
colnames(df)[colnames(df) %in% perc_4_all_2018] <- temp
perc_4_all_2018 <- temp
```

```{r q3 missing, results = 'hide'}
#Q3 arts supervisor
q3 <- q_arts %>% .[grepl("^Q3_", .)]
df %>% dplyr::select(q3) %>% sapply(function(x) sum(is.na(x)))
```

```{r q2 no designated supervisor, results = 'hide'}
df$Q2_1 %>% sum()
df$Q2_1 %>% mean()
```

### Arts Supervisors

New York City public schools were asked whether their arts supervisor was employed full- or part-time. If their supervisor was full-time, schools clarified whether they were solely working on arts programs or had other responsibilities.  
In small schools, or under-resourced ones, faculty may be expected to wear many hats. In the simplest cases, physical education teachers lead gym and health classes. More extreme examples can have teachers with certification in, for example, English teaching classes in math or science. I do not assert that an arts supervisor *must* be employed full-time to run an effective program, but I am curious as to what features predict their employment status.  

```{r q3 bar}
superv_status <- factor(c('Full-time, solely for arts', 'Full-time, with other duties', 'Part-time', 'None'),
                          levels=c('Full-time, solely for arts', 'Full-time, with other duties', 'Part-time', 'None'))

df %>% dplyr::select(q3) %>%
  colSums() %>% as_data_frame() %>%
  ggplot(aes(x=superv_status, y=value)) +
  geom_bar(stat = 'identity') +
  ggtitle('Arts Supervisor Employment Status') +
  xlab('Employment Status') + ylab('Frequency') +
  theme_eric() +
  theme(text = element_text(size=16),
        axis.text.x = element_text(size=8))
```

There are many full-time supervisors with duties other than the arts. I would be curious to discover the share of their responsibilities that are considered "other". It is possible that some administrators consider teaching an arts class or doing clerical work to be "other". Relative to the number of supervisors working full-time, there are few part-timers.  

```{r missing q3, results = 'hide'}
df$Q3 <- with(df, ifelse(Q3_1 == 1, '1',
                ifelse(Q3_2 == 1, '2',
                       ifelse(Q3_3 == 1, '3',
                              ifelse(Q3_4 == 1, '4', '0')))))
df$Q3 <- df$Q3 %>% factor(levels=seq(0,4))
df %>% filter(Q3 == '0') %>% count()
#Two schools did not respond to Q3
```

Only two schools did not respond to this question about their arts supervisor's status. In the previous [part](https://ecboxer.github.io/eda/art_education/arts_survey_1.html), I looked at the question of whether a school had a designated arts liaison. 72 schools responded that they did not, while 296 schools do not have any arts supervisor. I wonder whether the presence of an arts supervisor has more or less influence on student academic performance, relative to arts liaisons.  

To begin exploring academic performance, I will use the percentage of students to perform at or above standards for English Language Arts (ELA) and math.  

```{r math and ela scores, fig.width=8, fig.height=8}
df %>% dplyr::select(perc_34_all_2018, Q3) %>%
  filter(!is.na(perc_34_all_2018_ela)) %>% 
  filter(Q3 != 0) %>% 
  mutate(Q3 = factor(ifelse(Q3 == 1, 'Full-time, solely for arts',
                            ifelse(Q3 == 2, 'Full-time, with other duties',
                                   ifelse(Q3 == 3, 'Part-time', 'None'))))) %>% 
  ggplot(aes(x=perc_34_all_2018_ela, y=perc_34_all_2018_math)) +
  geom_point(alpha = .8) + facet_wrap(~Q3,
                            labeller = label_value) +
  ggtitle('Academic Performance, by Supervisor Status') +
  xlab('Students Passing, ELA (%)') +
  ylab('Students Passing, Math (%)') + theme_eric() +
  theme(text = element_text(size=16),
        axis.text.x = element_text(size=8))
```

I came into this analysis with some suspicion that arts programs might be more beneficial for ELA performance than for math, if they were to have any effect. In the plot above there is a positive relationship between ELA and math scores, regardless of art supervisor status.  

```{r supervisor status stats, results = 'hide'}
for (i in seq(4)) {
  lm.fit <- df %>% select(perc_34_all_2018, Q3) %>% 
    filter(Q3 == i) %>% 
    lm(perc_34_all_2018_ela~perc_34_all_2018_math, data=.)
  print(i)
  print(lm.fit %>% summary())
  print(lm.fit %>% confint())
}
```

```{r supervisor performance stats null}
lm.fit.null <- df %>% select(perc_34_all_2018) %>% 
  rename(Math_score = perc_34_all_2018_math,
         ELA_score = perc_34_all_2018_ela) %>% 
  lm(Math_score~., data=.)
lm.fit.null.summ <- lm.fit.null %>% summary()
lm.fit.null.summ
```
#### Output 1

On inspection the relation between the two scores seems identical between supervisor statuses. In fact, fitting a linear regression to each yields coefficients close to the uncontrolled coefficient for math scores on ELA scores, 1.06415 (Output 1). $Math\_score = \beta_0 + \beta_1 * ELA\_score$  

```{r supervisor performance stats control}
lm.fit <- df %>% select(q3, perc_34_all_2018) %>% 
  rename(Math_score = perc_34_all_2018_math,
         ELA_score = perc_34_all_2018_ela) %>% 
  lm(Math_score~., data=.)
lm.fit.summ <- lm.fit %>% summary()
lm.fit.summ
```
#### Output 2

We can control for supervisor status with dummy variables by altering the regression like so, $Math\_score = \beta_0 + \beta_1 * ELA\_score + \beta_2 * Q3\_1 + \beta_3 * Q3\_2 + \beta_4 * Q3\_3 + \beta_5 * Q3\_4$, where $Q3\_i$ corresponds to the $i^{th}$ member of the list `c('Full-time, solely for arts', 'Full-time, with other duties', 'Part-time', 'None')` (Output 2).  

```{r supervisor performance stats control confint}
lm.fit %>% confint()
```
#### Output 3

This model specification does not result in a compellingly significant coefficient for the effect of a particular supervisor status. Any status is estimated to have a positive coefficient, but a 95% confidence interval (Output 3) does not exclude the possibility that any of status coefficients could be zero.  

```{r supervisor performance diff, echo=TRUE}
delta.beta1 <- coefficients(lm.fit.null.summ)[2,1] - coefficients(lm.fit.summ)[6,1]
delta.beta1
se.delta.beta1 <- sqrt(coefficients(lm.fit.null.summ)[2,2]^2 + coefficients(lm.fit.summ)[6,2]^2)
se.delta.beta1
```
#### Output 4

On inspection, the regression coefficients for `ELA_score` are close between model specifications A (no controls) and B (controlling for supervisor status). The difference in estimated coefficients $\beta_1$ is -0.0003 (Output 4). To find the variability of the difference in regression coefficients, we use the formula $Var(A-B) = Var(A) + Var(B) - 2*Cov(A,B)$. Assuming covariance between the two estimates is zero, we arrive at 0.0201 as the standard error of the difference.  
Now I can assert that there is no difference in the relationsip between ELA and math scores from supervisor statuses.  

```{r mosaic, fig.width=15, fig.height=10}
library(ggmosaic)
df %>% select(Q2_1, Q3) %>% 
  mutate(Q2_1 = ifelse(Q2_1 == 1, 'No', 'Yes'),
         Q3 = factor(ifelse(Q3 == 1, 'Full-time, solely for arts',
                            ifelse(Q3 == 2, 'Full-time, with other duties',
                                   ifelse(Q3 == 3, 'Part-time', 'None'))))) %>% 
  mutate(Q3 = as_factor(Q3, levels = c('Full-time, solely for arts',
                    'Full-time, with other duties',
                    'Part-time',
                    'None'))) %>% 
  ggplot() + 
  geom_mosaic(aes(x=product(Q3, Q2_1), fill=Q3), na.rm=T) +
  scale_fill_manual(values=c5) +
  ggtitle('Does Having an Arts Liaison Affect Supervisor Status?') +
  xlab('School arts liaison?') +
  ylab('Arts Supervisor Status') + theme_eric() +
  theme(text = element_text(size=20),
        axis.text.x = element_text(size=12))
```

This mosaic plot illustrates that the proportion of schools without an arts program supervisor is greater for schools without an arts liaison that for those with. This lends support to the idea that those schools are lacking the resources to fully staff their arts programs, as they have not filled two key positions. It is of course possible that arts liaisons and supervisors are not necessary to effective programs, and schools without either are running just fine. I would like to assess the quality of the arts programs themselves as a function their liaison and supervisor statuses, perhaps through some sort of measure of funding or arts resources.  

***

The next questions concern certifications that arts supervisors may have, either in an arts discipline or in administration.  

```{r supervisor cert both, results = 'hide'}
df %>% filter(Q4_1 == 1 & Q5_1 == 1) %>% count()
df %>% filter(Q4_1 == 1) %>% count()
df %>% filter(Q5_1 == 1) %>% count()
```

```{r supervisor cert}
df$supervisor_cert <- with(df, ifelse(Q4_1 == 1 & Q5_1 == 1, 'Both',
                ifelse(Q4_1 == 1 & Q5_1 == 0, 'Arts',
                       ifelse(Q4_1 == 0 & Q5_1 == 1, 'Administration', 'None'))))
df %>% dplyr::select(Q0_DBN, supervisor_cert) %>% 
  group_by(supervisor_cert) %>% 
  count() %>% 
  ggplot(aes(x=supervisor_cert, y=n)) +
  geom_bar(stat='identity') +
  ggtitle('Does the Arts Supervisor Have Any Certifications?') +
  xlab('Supervisor Certification') + ylab('Frequency') +
  theme(text = element_text(size=16),
        axis.text.x = element_text(size=8))
```

Few supervisors are certified in the arts, the majority are administrators. Could this have an impact on the efficacy of an arts program?  

```{r supervisor cert stats, results='hide'}
df <- df %>%
  mutate(supervisor_cert_both = ifelse(Q4_1 == 1 & Q5_1 == 1, 1, 0))
q_certs <- c('Q4_1', 'Q5_1', 'supervisor_cert_both')
df %>%
  dplyr::select(q_certs, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
df %>%
  dplyr::select(q_certs, perc_34_all_2018_math) %>% 
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
```

Linear regressions do not yield a statistically significant result for the effect of either or both supervisor certification on student academic performance. The model specifications I tried out were of the form $score = \beta_0 + \beta_1 * arts\_cert + \beta_2 * admin\_cert + \beta_3 * both\_cert$.  

```{r supervisor cert stats ctrl}
df <- df %>%
  mutate(supervisor_cert_both = ifelse(Q4_1 == 1 & Q5_1 == 1, 1, 0))
q_certs <- c('Q4_1', 'Q5_1', 'supervisor_cert_both')
lm.ela <- df %>%
  dplyr::select(q_certs, total_enrollment_2017, perc_black_2017, perc_pov_2017, perc_34_all_2018_ela) %>% 
  rename(arts_cert = Q4_1,
         admin_cert = Q5_1,
         both_cert = supervisor_cert_both) %>% 
  lm(perc_34_all_2018_ela~., data=.)
lm.ela %>% summary()
lm.ela %>% confint()
lm.math <- df %>%
  dplyr::select(q_certs, total_enrollment_2017, perc_black_2017, perc_pov_2017, perc_34_all_2018_math) %>% 
  rename(arts_cert = Q4_1,
         admin_cert = Q5_1,
         both_cert = supervisor_cert_both) %>% 
  lm(perc_34_all_2018_math~., data=.)
lm.math %>% summary()
lm.math %>% confint()
```
#### Output 5

After controlling for school size and percentage of black and impoverished students, there is no statistically significant effect for certifications on academic performance. A 95% confidence interval of the coefficient for supervisor certification in administration includes zero for ELA and math scores, so we cannot claim that there is a statistically significant nonzero effect on academic performance. The confidence interval for arts or both certifications are even wider.  

```{r supervisor cert stats narrow ctrl}
df <- df %>%
  mutate(supervisor_cert_both = ifelse(Q4_1 == 1 & Q5_1 == 1, 1, 0))
q_certs <- c('Q4_1', 'Q5_1', 'supervisor_cert_both')
lm.ela <- df %>%
  dplyr::select(q_certs, total_enrollment_2017, perc_black_2017, perc_pov_2017, perc_4_all_2018_ela) %>% 
  rename(arts_cert = Q4_1,
         admin_cert = Q5_1,
         both_cert = supervisor_cert_both) %>% 
  lm(perc_4_all_2018_ela~., data=.)
lm.ela %>% summary()
lm.ela %>% confint()
lm.math <- df %>%
  dplyr::select(q_certs, total_enrollment_2017, perc_black_2017, perc_pov_2017, perc_4_all_2018_math) %>% 
  rename(arts_cert = Q4_1,
         admin_cert = Q5_1,
         both_cert = supervisor_cert_both) %>% 
  lm(perc_4_all_2018_math~., data=.)
lm.math %>% summary()
lm.math %>% confint()
```
#### Output 6

If we drill down further, looking at only the percentage of students to receive a 4 (the highest grade), then there is a statistically significant coefficient for the effect of both certifications on math scores, with a p-value of 0.048996 (Output 6). In this case, a 95% confidence interval just excludes zero, being [0.025, 11.354].  
Having been able to arrive at this more tenuous result I feel a renewed belief that arts programs do have an effect on academic performance and that some proof is lying in the data somewhere. On the other hand I am suspicious as to how I *manufactured* this result by narrowing my focus until I hacked my way to an accceptable p-value.  

## Data Provenance
The [2017-2018 Arts Survey Data](https://data.cityofnewyork.us/Education/2017-2018-Arts-Survey-Data/475h-cg5t) has data about arts teachers, budgets, partnerships with cultural organizations and parental involvement in NYC public schools.  

In an effort to gain greater context for this data, I have examined it in conjuction with publicly available [ELA and Math state test results](https://infohub.nyced.org/reports-and-policies/citywide-information-and-data/test-results) and [demographic data](https://data.cityofnewyork.us/Education/2013-2018-Demographic-Snapshot-School/s52a-8aq6).  

My goals are to understand the state of arts programs in NYC schools, what variables affect the resources of arts programs, and whether arts programs have an effect on the academic performance of students.  
