---
author: "Eric Boxer"
date: "January 27, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

```{r load data}
library(dagitty)
library(ggplot2)
library(ggpubr)
library(ggridges)
library(MatchIt)
library(tidyverse)
df <- read_csv('resources/arts_demo_ela_math.csv')
theme_eric <- function(base_size = 11,
                      base_family = 'URWHelvetica',
                      base_line_size = base_size / 170,
                      base_rect_size = base_size / 170){
  theme_bw(base_size = base_size, 
                base_family = base_family,
                base_line_size = base_line_size) %+replace%
    theme(
      plot.title = element_text(
        color = rgb(25, 43, 65, maxColorValue = 255), 
        face = "bold",
        hjust = 0),
      axis.title = element_text(
        color = rgb(105, 105, 105, maxColorValue = 255),
        size = rel(0.75)),
      axis.text = element_text(
        color = rgb(105, 105, 105, maxColorValue = 255),
        size = rel(0.5)),
      panel.grid.major = element_line(
        rgb(105, 105, 105, maxColorValue = 255),
        linetype = "dotted"),   
      panel.grid.minor = element_line(
        rgb(105, 105, 105, maxColorValue = 255),
        linetype = "dotted", 
        size = rel(4)),
      
      complete = TRUE
    )
}
theme_set(theme_eric())
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
c1 <- '#000044'
c4 <- c('#222222', '#555555', '#888888', '#BBBBBB')
c5 <- c('#222222', '#444444', '#666666', '#888888', '#AAAAAA')
```

```{r q names}
q_arts <- df %>% colnames() %>% .[grepl("^Q", .)]
q_ela <- df %>% colnames() %>% .[grepl("ela$", .)]
q_math <- df %>% colnames() %>% .[grepl("math$", .)]
q_demo <- df %>% colnames() %>% setdiff(union(q_arts, union(q_ela, q_math)))
disc_4 <- c("Dance", "Music", "Theater", "Visual Arts")
disc_5 <- c("Dance", "Film", "Music", "Theater", "Visual Arts")
```

```{r get perc_34, results = 'hide'}
#Get performance metrics of interest, % of passing students
perc_34_ela <- q_ela %>% .[grepl("^perc_34", .)]
perc_34_math <- q_math %>% .[grepl("^perc_34", .)]

#Get percent passing for all grades in 2018.
perc_34_all_2018 <- union(perc_34_ela, perc_34_math) %>%
  .[grepl("All Grades_2018", .)]
df %>% dplyr::select(perc_34_all_2018) %>% is.na() %>% colSums()
#Missing data for 370 schools
#Get percent passing for 4th graders in 2018
perc_34_4_2018 <- union(perc_34_ela, perc_34_math) %>% 
  .[grepl("4_2018", .)]
df %>% dplyr::select(perc_34_4_2018) %>% is.na() %>% colSums()

#Rename columns
temp <- c('perc_34_all_2018_ela', 'perc_34_all_2018_math')
colnames(df)[colnames(df) %in% perc_34_all_2018] <- temp
perc_34_all_2018 <- temp

perc_4_all_2018 <- df %>% colnames() %>% .[grepl('perc_4_', .)] %>% 
  .[grepl('_All Grades_2018', .)]
#Rename perc_4
temp <- c('perc_4_all_2018_ela', 'perc_4_all_2018_math')
colnames(df)[colnames(df) %in% perc_4_all_2018] <- temp
perc_4_all_2018 <- temp
```

### Resources devoted to arts education

This is part of an extended look at the NYC School Arts Survey:  
[Part One: Arts Education Liaisons](https://ecboxer.github.io/eda/art_education/arts_survey_1.html)  
[Part Two: Arts Education Supervisors](https://ecboxer.github.io/eda/art_education/arts_survey_2.html)  

The survey had several questions regarding which resources schools are devoting to arts education, whether the administrator thought they were sufficient and which non-Department of Education sources of support were available to them.    

Schools were asked how many rooms they have dedicated to arts education. Rooms were divided into two categories: "rooms designed and used solely for the arts" and "multi-purpose or general education classroooms used for arts education". Five arts disciplines were considered: dance, music, theater, visual arts, and media arts (a category which includes film and photography). Note that the term media arts refers to film programs.  
Number of rooms is certainly not a direct metric for a school's commitment to the arts. I hypothesize that it may prove useful for assessing the resources that are made available for arts education more broadly. Of course, there may be confounding factors, such as school size and borough.  

```{r rooms setup}
q8 <- df %>% colnames() %>% .[grepl('Q8_', .)]
df <- df %>%
  mutate(dance_rooms = Q8_R1_C1 + Q8_R1_C2,
         music_rooms = Q8_R2_C1 + Q8_R2_C2,
         theater_rooms = Q8_R3_C1 + Q8_R3_C2,
         visual_arts_rooms = Q8_R4_C1 + Q8_R4_C2,
         media_arts_rooms = Q8_R5_C1 + Q8_R5_C2)
rooms <- df %>% colnames %>% .[grepl('rooms', .)]
df %>% 
  dplyr::select(ends_with('rooms')) %>% 
  gather(key = discipline,
         value = num) %>%
  ggplot(aes(x=num)) +
  geom_density() +
  facet_wrap(~discipline, labeller=label_value) +
  xlab('Number of rooms') +
  ggtitle('Number of rooms used in arts education') +
  theme_eric()
df %>% 
  dplyr::select(ends_with('rooms')) %>% 
  gather(key = discipline,
         value = num) %>%
  ggplot(aes(x=num)) +
  geom_histogram(bins=30) +
  facet_wrap(~discipline, labeller=label_value) +
  xlab('Number of rooms') + 
  theme_eric()
```

All disciplines show the same trend; many schools have zero to two rooms for a given discipline and a few schools have more than twenty rooms. Having twenty-some rooms might be bordering on unrealistic, but the above plots depict both categories of rooms, those designed solely for the arts and multi-purpose. Will we get more reasonable numbers by examining the room counts separately?  

```{r rooms solely}
df %>% select(q8) %>%
  select(ends_with('C1')) %>% 
  mutate(dance_rooms = Q8_R1_C1,
         music_rooms = Q8_R2_C1,
         theater_rooms = Q8_R3_C1,
         visual_arts_rooms = Q8_R4_C1,
         media_arts_rooms = Q8_R5_C1) %>% 
  select(ends_with('rooms')) %>% 
  gather(key = discipline,
         value = num) %>% 
  filter(!is.na(num)) %>% 
  ggplot(aes(x=num)) +
  geom_density() +
  facet_wrap(~discipline) +
  ggtitle('Rooms designed and used solely for the arts') +
  theme_eric()
df %>% select(q8) %>%
  select(ends_with('C2')) %>% 
  mutate(dance_rooms = Q8_R1_C2,
         music_rooms = Q8_R2_C2,
         theater_rooms = Q8_R3_C2,
         visual_arts_rooms = Q8_R4_C2,
         media_arts_rooms = Q8_R5_C2) %>% 
  select(ends_with('rooms')) %>% 
  gather(key = discipline,
         value = num) %>% 
  filter(!is.na(num)) %>% 
  ggplot(aes(x=num)) +
  geom_density() +
  facet_wrap(~discipline) +
  ggtitle('Multi-purpose and general education classrooms used for arts education') + 
  theme_eric()
```

Looking at the number of rooms designed solely for the arts, the maximum number of rooms is a more reasonable thirteen. For a school with several hundred students and a dedicated arts program I can picture that. If we look at multi-purpose rooms, we see that this class contains the bulk of the rooms. Not many schools have dozens of multi-purpose rooms for arts disciplines, but again, in the context of a large school, it seems like a possible number for some schools to have.  
I would be curious to discover the overlap for multi-purpose rooms among arts disciplines. In other words, is a school with twenty multi-purpose rooms reporting some of those rooms as in use for multiple arts discipline? I do not see any way to completely correct for that possibility, but it would be interesting to control for school size to try to get to the bottom of it. If this does not work, then we could consider only the number of rooms designed solely for art, but this could penalize small schools unduly.  

```{r rooms stats}
#ELA
df %>% dplyr::select(rooms, total_enrollment_2017, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
#Math
df %>% dplyr::select(rooms, total_enrollment_2017, perc_34_all_2018_math) %>%
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
```

Conditioning on the effect of school size, through total enrollment, there is no statistically significant effect for the total number of rooms for arts education on student academic performance.  

```{r rooms stats ctrl}
rooms_ded <- df %>% select(q8) %>%
  dplyr::select(ends_with('C1')) %>% colnames()
rooms_mul <- df %>% select(q8) %>%
  dplyr::select(ends_with('C2')) %>% colnames()
#ELA dedicated
df %>% dplyr::select(rooms_ded, total_enrollment_2017, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
#Math dedicated
df %>% dplyr::select(rooms_ded, total_enrollment_2017, perc_34_all_2018_math) %>%
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
```

```{r rooms stats ctrl multi, results='hide'}
#ELA multi
df %>% dplyr::select(rooms_mul, total_enrollment_2017, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
#Math multi
df %>% dplyr::select(rooms_mul, total_enrollment_2017, perc_34_all_2018_math) %>%
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
```

Controlling for school size, through total student enrollment, we can see a statistically significant (p-value less than 0.001) positive coefficient for the effect of additional rooms designed and used solely for music, on ELA state test scores. If we consider math scores, rooms solely dedicated to music still have a positive effect, but at a 0.0044 p-value. No other arts disciplines have a statistically significant effect on ELA scores, but dance and media arts rooms have statistically significant (p-value less than 0.05) negative effects on math scores.  
When we consider multi-purpose rooms, no arts discipline has a statistically significant effect on academic performance.  
I am curious about the reason behind why music rooms appear to be associated with higher test scores, and why dance and media arts are associated with lower scores. To investigate we can try to control for different features and divide the performance metric by grade.  

```{r rooms grades}
ela_2018 <- q_ela %>% .[grepl('perc_34_[3-8]_2018',.)]
math_2018 <- q_math %>% .[grepl('perc_34_[3-8]_2018',.)]
for (i in ela_2018) {
  temp <- df %>%
    dplyr::select(i, rooms_ded, total_enrollment_2017) %>% 
    .[complete.cases(.),]
  f.ela.1 <- paste(as.character(i), 'total_enrollment_2017', sep='~')
  f.ela.2 <-  paste(rooms_ded, collapse='+')
  f.ela <- as.formula(paste(f.ela.1, f.ela.2, sep='+'))
  temp <- temp %>% lm(f.ela, data=.)
  if (i %in% c('perc_34_5_2018_ela', 'perc_34_6_2018_ela')){
    print(i)
    print(temp %>% summary())
    # print(temp %>% confint.default())
  }
}
for (i in math_2018) {
  temp <- df %>%
    dplyr::select(i, rooms_ded, total_enrollment_2017) %>% 
    .[complete.cases(.),]
  f.math.1 <- paste(as.character(i), 'total_enrollment_2017', sep='~')
  f.math.2 <-  paste(rooms_ded, collapse='+')
  f.math <- as.formula(paste(f.math.1, f.math.2, sep='+'))
  temp <- temp %>% lm(f.math, data=.)
  if (i %in% c('perc_34_5_2018_math', 'perc_34_6_2018_math')){
    print(i)
    print(temp %>% summary())
    # print(temp %>% confint.default())
  }
}
```

If we break down academic performance by grade, then the results are not as straigtforward. For sixth-, seventh- and eighth-graders' ELA performance, media arts rooms have a more statistically significant negative coefficient than the positive coefficient for music rooms. All grades show a positive coefficient for music rooms, with a p-value no more than 0.01. There is a large jump in media arts between fifth- and sixth-grades, which is the demarcation between elementary and middle schools. A jump of this nature could have an underlying reason, beyond increased susceptibility to media arts programs at the expense of academic performance beginning in middle school. Perhaps not many elementary schools have media arts programs at all, or they are more commonly found in schools that are otherwise performing at an atypical level (higher or lower). Both of these can be checked.  
Math scores show a similar trend to ELA scores, with most grades showing a statistically significant positive coefficient for music rooms. Among middle schools, media arts rooms have a statistically significant negative coefficient. The key difference with ELA scores is that there is not a statistically significant positive coefficient for music rooms on eighth-grade math scores.  

```{r media arts elem}
q_elem <- q_demo %>% .[grepl('grd_[1-5]_2017', .)]
q_mid <- q_demo %>% .[grepl('grd_[6-8]_2017', .)]
disc_5.1 <- c('Dance', 'Music', 'Theater', 'Visual Arts', 'Media Arts')
for (i in 1:5){
  temp <- df %>% select(q_elem[i], rooms_ded) %>% 
    filter((!!sym(q_elem[i])) > 0) %>% 
    select(-(!!sym(q_elem[i]))) %>% 
    colSums(na.rm = T) %>%
    as_data_frame() %>% 
    ggplot(aes(x=disc_5.1, y=value)) +
    geom_bar(stat='identity') +
    ggtitle(q_elem[i]) + theme_eric()
  if (i == 5) {
    print(temp)
  }
}
for (i in 1:3){
  temp <- df %>% select(q_mid[i], rooms_ded) %>% 
    filter((!!sym(q_mid[i])) > 0) %>% 
    select(-(!!sym(q_mid[i]))) %>% 
    colSums(na.rm = T) %>%
    as_data_frame() %>% 
    ggplot(aes(x=disc_5.1, y=value)) +
    geom_bar(stat='identity') +
    ggtitle(q_mid[i]) + theme_eric()
  if (i == 1) {
    print(temp)
  }
}
for (i in q_elem){
  temp <- df %>% select(i, rooms_ded) %>% 
    filter((!!sym(i)) > 0) %>% 
    select(-(!!sym(i))) %>% 
    mutate(dance_rooms = Q8_R1_C1,
           music_rooms = Q8_R2_C1,
           theater_rooms = Q8_R3_C1,
           visual_arts_rooms = Q8_R4_C1,
           media_arts_rooms = Q8_R5_C1) %>% 
    select(ends_with('rooms')) %>%
    gather(key = discipline,
           value = num) %>% 
    ggplot(aes(x=num)) +
    geom_density() +
    facet_wrap(~discipline) +
    ggtitle(i) + theme_eric()
  if (i == 'grd_5_2017') {
    print(temp)
  }
}
for (i in q_mid){
  temp <- df %>% select(i, rooms_ded) %>% 
    filter((!!sym(i)) > 0) %>% 
    select(-(!!sym(i))) %>% 
    mutate(dance_rooms = Q8_R1_C1,
           music_rooms = Q8_R2_C1,
           theater_rooms = Q8_R3_C1,
           visual_arts_rooms = Q8_R4_C1,
           media_arts_rooms = Q8_R5_C1) %>% 
    select(ends_with('rooms')) %>%
    gather(key = discipline,
           value = num) %>% 
    ggplot(aes(x=num)) +
    geom_density() +
    facet_wrap(~discipline) +
    ggtitle(i) + theme_eric()
  if (i == 'grd_6_2017') {
    print(temp)
  }
}
```

I do not see a significant difference in the distribution of art rooms between elementary and middle schools. Music and visual arts rooms are the most common. Let's look at the performance of schools with media arts programs.  
We can match schools on having media arts rooms while controlling for number of students, number of students in each grade, and perhaps other demographic and academic features. Then we can examine the effect of media arts rooms on academic performance.  

```{r rooms media match}
library(MatchIt)
#Feature for having rooms dedicated to media arts
df <- df %>% filter(!is.na(Q8_R5_C1)) %>% 
  mutate(rm_ded_media = as.integer(pmin(1, Q8_R5_C1, na.rm=T)))
f.1 <- paste(paste(q_elem, collapse='+'), paste(q_mid, collapse='+'), sep='+')
f <- as.formula(paste('rm_ded_media~perc_34_all_2018_ela+total_enrollment_2017', f.1, sep='+'))
temp <- df %>%
  dplyr::select(rm_ded_media, perc_34_all_2018_ela, total_enrollment_2017, q_elem, q_mid) %>% 
  .[complete.cases(.),]
m.out1 <- matchit(f, data=temp, method='nearest', distance='logit')
summary(m.out1)
plot(m.out1)
```

Propensity score matching improves balance for most features.  

```{r rooms ps match}
m.data1 <- match.data(m.out1, distance='pscore')
# hist(m.data1$pscore, breaks=30)
# summary(m.data1$pscore)

with(m.data1, t.test(perc_34_all_2018_ela ~ rm_ded_media))

lm.treat1 <- lm(perc_34_all_2018_ela~rm_ded_media, data=m.data1)
summary(lm.treat1)

lm.treat2 <- m.data1 %>% select(-weights) %>%
  lm(perc_34_all_2018_ela~., .)
summary(lm.treat2)
```

Matching suggests that rooms dedicated to media arts are associated with schools with higher ELA scores. The result that led to this investigation was that rooms dedicated to media arts are associated with lower state test scores among middle-schoolers. In light of this analysis I cannot draw any firm conclusions, but would be interested to find some instrumental variable to use in lieu of an experiment.  

## NYC Schools Arts Survey Data
The [2017-2018 Arts Survey Data](https://data.cityofnewyork.us/Education/2017-2018-Arts-Survey-Data/475h-cg5t) has data about arts teachers, budgets, partnerships with cultural organizations and parental involvement in NYC public schools.  

In an effort to gain greater context for this data, we can examine it in conjuction with publicly available [ELA and Math state test results](https://infohub.nyced.org/reports-and-policies/citywide-information-and-data/test-results) and [demographic data](https://data.cityofnewyork.us/Education/2013-2018-Demographic-Snapshot-School/s52a-8aq6).  

My goals were to understand the state of arts programs in NYC schools, what variables affect the resources of arts programs, and whether arts programs have an effect on the academic performance of students.  
