---
author: "Eric Boxer"
date: "January 17, 2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      cache = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

## NYC Schools Arts Survey Data
The [2017-2018 Arts Survey Data](https://data.cityofnewyork.us/Education/2017-2018-Arts-Survey-Data/475h-cg5t) has data about arts teachers, budgets, partnerships with cultural organizations and parental involvement in NYC public schools.  

In an effort to gain greater context for this data, we can examine it in conjuction with publicly available [ELA and Math state test results](https://infohub.nyced.org/reports-and-policies/citywide-information-and-data/test-results) and [demographic data](https://data.cityofnewyork.us/Education/2013-2018-Demographic-Snapshot-School/s52a-8aq6).  

My goals were to understand the state of arts programs in NYC schools, what variables affect the resources of arts programs, and whether arts programs have an effect on the academic performance of students.  

```{r load data}
library(dagitty)
library(ggplot2)
library(ggpubr)
library(ggridges)
library(MatchIt)
library(tidyverse)
df <- read_csv('resources/arts_demo_ela_math.csv')
theme_eric <- function(base_size = 11,
                      base_family = 'URWHelvetica',
                      base_line_size = base_size / 170,
                      base_rect_size = base_size / 170){
  theme_bw(base_size = base_size, 
                base_family = base_family,
                base_line_size = base_line_size) %+replace%
    theme(
      plot.title = element_text(
        color = rgb(25, 43, 65, maxColorValue = 255), 
        face = "bold",
        hjust = 0),
      axis.title = element_text(
        color = rgb(105, 105, 105, maxColorValue = 255),
        size = rel(0.75)),
      axis.text = element_text(
        color = rgb(105, 105, 105, maxColorValue = 255),
        size = rel(0.5)),
      panel.grid.major = element_line(
        rgb(105, 105, 105, maxColorValue = 255),
        linetype = "dotted"),   
      panel.grid.minor = element_line(
        rgb(105, 105, 105, maxColorValue = 255),
        linetype = "dotted", 
        size = rel(4)),
      
      complete = TRUE
    )
}
theme_set(theme_eric())
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
c1 <- '#000044'
c4 <- c('#222222', '#555555', '#888888', '#BBBBBB')
c5 <- c('#222222', '#444444', '#666666', '#888888', '#AAAAAA')
```

```{r q names}
q_arts <- df %>% colnames() %>% .[grepl("^Q", .)]
q_ela <- df %>% colnames() %>% .[grepl("ela$", .)]
q_math <- df %>% colnames() %>% .[grepl("math$", .)]
q_demo <- df %>% colnames() %>% setdiff(union(q_arts, union(q_ela, q_math)))
disc_4 <- c("Dance", "Music", "Theater", "Visual Arts")
disc_5 <- c("Dance", "Film", "Music", "Theater", "Visual Arts")
```

```{r get perc_34, results = 'hide'}
#Get performance metrics of interest, % of passing students
perc_34_ela <- q_ela %>% .[grepl("^perc_34", .)]
perc_34_math <- q_math %>% .[grepl("^perc_34", .)]

#Get percent passing for all grades in 2018.
perc_34_all_2018 <- union(perc_34_ela, perc_34_math) %>%
  .[grepl("All Grades_2018", .)]
df %>% dplyr::select(perc_34_all_2018) %>% is.na() %>% colSums()
#Missing data for 370 schools
#Get percent passing for 4th graders in 2018
perc_34_4_2018 <- union(perc_34_ela, perc_34_math) %>% 
  .[grepl("4_2018", .)]
df %>% dplyr::select(perc_34_4_2018) %>% is.na() %>% colSums()

#Rename columns
temp <- c('perc_34_all_2018_ela', 'perc_34_all_2018_math')
colnames(df)[colnames(df) %in% perc_34_all_2018] <- temp
perc_34_all_2018 <- temp
```

```{r q3 missing, results = 'hide'}
#Q3 arts supervisor
q3 <- q_arts %>% .[grepl("^Q3_", .)]
df %>% dplyr::select(q3) %>% sapply(function(x) sum(is.na(x)))
```

```{r q2 no designated supervisor, results = 'hide'}
df$Q2_1 %>% sum()
df$Q2_1 %>% mean()
```

### Arts Education Liaison
In 2018, 1265 schools were surveyed. Schools were asked whether they have a designated arts supervisor. 5.6%, 72 schools, had no arts supervisor. My first question has to do with whether there is any relation between the presence of an arts supervisor and the demographics of the student population.

```{r supervisor and demos}
df$Q2_factor <- df$Q2_1 %>% as.factor()
g1 <- df %>% dplyr::select(Q2_factor, total_enrollment_2017) %>% 
  ggplot(aes(x = total_enrollment_2017,
             color = Q2_factor,
             stat(density))) +
  geom_freqpoly(alpha = .5, size = 1) +
  scale_color_manual(values = c4[c(1,3)],
                    labels = c('No', 'Yes'),
                    name = 'Arts Supervisor?') +
  ylab('') + xlab('Total Enrollment')
g2 <- df %>% dplyr::select(Q2_factor, perc_pov_2017) %>% 
  ggplot(aes(x = perc_pov_2017,
             color = Q2_factor,
             stat(density))) +
  geom_freqpoly(alpha = .5, size = 1) +
  scale_color_manual(values = c4[c(1,3)],
                    labels = c('No', 'Yes'),
                    name = 'Arts Supervisor?') +
  ylab('') + xlab('Student Poverty (%)')
g3 <- df %>% dplyr::select(Q2_factor, perc_female_2017) %>% 
  ggplot(aes(x = perc_female_2017,
             color = Q2_factor,
             stat(density))) +
  geom_freqpoly(alpha = .5, size = 1) +
  scale_color_manual(values = c4[c(1,3)],
                    labels = c('No', 'Yes'),
                    name = 'Arts Supervisor?') +
  ylab('') + xlab('Female Students (%)')
g4 <- df %>% dplyr::select(Q2_factor, perc_black_2017) %>% 
  ggplot(aes(x = perc_black_2017,
             color = Q2_factor,
             stat(density))) +
  geom_freqpoly(alpha = .5, size = 1) +
  scale_color_manual(values = c4[c(1,3)],
                    labels = c('No', 'Yes'),
                    name = 'Arts Supervisor?') +
  ylab('') + xlab('Black Students (%)')
figure <- ggarrange(g1, g2, g3, g4,
                    labels = c('1', '2', '3', '4'),
                    ncol = 2,
                    nrow = 2)
figure
```

Schools without a designated arts supervisor appear to have more black students than those with supervisors (Fig 4). There appear to be several schools withan arts supervisor and very high poverty, but there is no clear relation (Fig 2). I also do not see a significant relation between total enrollment (Fig 1) or student gender ratio (Fig 3) and having an arts supervisor. Most schools have about 500 students and gender splits close to 50-50 so a small difference would be hard to detect in these graphs.  

In order to assess the significance of the differences pointed we see above, we can try to fit a logistic regression to the relationship between each variable and the binary response "Does the school have an an arts supervisor?"

```{r supervisor and demos stats, results = 'hide'}
demos <- c('total_enrollment_2017', 'perc_pov_2017', 'perc_female_2017', 'perc_black_2017')
rel_risk = list()
for (i in demos) {
  temp <- df %>% dplyr::select(Q2_1, i) %>% 
    glm(Q2_1~., data=., family='binomial')
  print(i)
  print(temp %>% summary())
  print(temp %>% confint.default())
  
  c.int <- temp$coefficients[1]
  c.i <- temp$coefficients[2]
  control <- exp(c.int) / (1 + exp(c.int))
  rel_risk[i] <- exp(c.i) / (1 - control + (control * exp(c.i)))
}
```

These regressions lend support to the differences spotted in the plots of student poverty and race, but the most statistically significant coefficient comes from total enrollment. The coefficient for total enrollment is negative with a p-value below a significance level of 0.001. Percentages of impoverished and black students have positive coefficients with p-values below a significance level of 0.01. As we had assumed from the graphs, there was no significant relation between gender and whether a school had an arts supervisor.  
In order to better interpret these results I converted the standard R logistic regression coefficients into relative risk, [as seen here](https://www.bmj.com/content/348/bmj.f7450.full?ijkey=NHT1YVsoX1RCm8r&keytype=ref).

```{r supervisor rel risk}
rel_risk[c(1, 2, 4)]
```

An increase in total enrollment of one student is associated with a .25% decrease in the likelihood that a school has an arts supervisor. An increase of one percentage point in student poverty and in share of black students is associated with 2.13% and 1.26% increases, respectively, in the likelihood that a school has an arts supervisor.

```{r supervisor and demos all, results='hide'}
df %>% dplyr::select(Q2_1, demos) %>% dplyr::select(-perc_female_2017) %>% 
  glm(Q2_1~., data=., family='binomial') %>% 
  summary()
df %>% dplyr::select(demos) %>% dplyr::select(-perc_female_2017) %>% 
  lm(total_enrollment_2017~., data=.) %>% 
  summary()
```

Consider the possibility that there is confounding between these variables, ie. small schools tend to serve poor, mostly black student populations. To address this possibility I ran a logistic regression upon total enrollment, student poverty and share of black students. This yielded only one significant feature, total enrollment, supporting some sort of statistical interdependence between the variables.  
Let S denote school art supervisor, E enrollment, P poverty and B black students.

```{r cg supervisor}
g <- dagitty('dag {
    S [pos="0,1"]
    E [pos="1,2"]
    P [pos="2,1"]
    B [pos="1,0"]
    
    E -> S
    P -> S
    B -> S
    P -> E
    B -> E
    B <-> P
}')
plot(g)
```

To further investigate confounding, I ran a linear model for total enrollment on percentage of impoverished and black students. As suspected, both had negative relationships with total enrollment. In particular, with a p-value less than significance level 0.001, each additional percentage point of black students was associated with a decrease of 5.22 students.  

As a former NYC public school student this coincides with my experience of the school system. In my old neighborhood of Canarsie in Brooklyn was a school previously known as South Shore High School. It served thousands of poor, majority minority students. In 2010, the school was converted into the [South Shore Educational Complex](https://insideschools.org/school/18K515) and now houses five smaller high schools. A naive observer could be expected to assume that small schools have higher faculty:student ratios and more resources to expend on their students. In reality the policies of the NYC DOE have led to the exact opposite, with underperforming schools split up without necessarily being given the resources to turn around their performance.  

My second question is whether there is any relation between having an arts supervisor and student academic performance.  
```{r supervisor performance}
df %>% select(perc_34_all_2018_ela, Q2_1) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>%
  summary()
df %>% select(perc_34_all_2018_ela, Q2_1, perc_pov_2017, perc_black_2017, total_enrollment_2017) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>%
  summary()
df %>% select(perc_34_all_2018_math, Q2_1) %>% 
  lm(perc_34_all_2018_math~., data=.) %>%
  summary()
df %>% select(perc_34_all_2018_math, Q2_1, perc_pov_2017, perc_black_2017, total_enrollment_2017) %>% 
  lm(perc_34_all_2018_math~., data=.) %>%
  summary()
```

To measure academic performance, we can use the percentage of students to receive a 3 or 4 on state standardized tests (grades corresponding to at or above expectations), denoted `perc_34`. If we consider the relationship between having an arts supervisor and academic performance, then we see an 11 and 14 point decrease in English Language Arts (ELA) and math, respectively, at a significance level less than 0.001.  
However, considering the causal relationship above, we can introduce context by conditioning on student poverty, share of black students, and total enrollment. In this case we do not have any statistically significant relation with ELA scores and a four point decrease in math scores, but at a 0.04 significance level.  

Next time, I will continue going through the survey. The presence of an arts supervisor does not necessarily equate to a well-resourced and efficacious arts program, so I hope to find some interesting questions.


```{r q3 bar}
superv_status <- factor(c('FT, solely arts', 'FT, other', 'PT', 'None'),
                          levels=c('FT, solely arts', 'FT, other', 'PT', 'None'))

df %>% dplyr::select(q3) %>%
  colSums() %>% as_data_frame() %>%
  ggplot(aes(x=superv_status, y=value)) +
  geom_bar(stat = 'identity')
```

There are many full-time supervisors with duties other than the arts. There are very few part-time supervisors of those working full-time and solely dedicated to the arts.

```{r missing q3, results = 'hide'}
df$Q3 <- with(df, ifelse(Q3_1 == 1, '1',
                ifelse(Q3_2 == 1, '2',
                       ifelse(Q3_3 == 1, '3',
                              ifelse(Q3_4 == 1, '4', '0')))))
df$Q3 <- df$Q3 %>% factor(levels=seq(0,4))
df %>% filter(Q3 == '0') %>% count()
#Two schools did not respond to Q3
```

```{r math and ela scores}
df %>% dplyr::select(perc_34_all_2018, Q3) %>%
  filter(!is.na(perc_34_all_2018_ela)) %>% 
  filter(Q3 != 0) %>% 
  ggplot(aes(x=perc_34_all_2018_ela, y=perc_34_all_2018_math)) +
  geom_point() + facet_wrap(vars(Q3))
```

There is a positive relationship between ELA and math scores, regardless of the status of an arts supervisor in the schools.

The next questions are in regard to certifications that the arts supervisor may have, in an arts discipline or in administration. Are there any supervisors certified in an arts discipline and in administration?

```{r supervisor cert both, results = 'hide'}
df %>% filter(Q4_1 == 1 & Q5_1 == 1) %>% count()
```

```{r supervisor cert}
df$supervisor_cert <- with(df, ifelse(Q4_1 == 1 & Q5_1 == 1, 'both',
                ifelse(Q4_1 == 1 & Q5_1 == 0, 'arts',
                       ifelse(Q4_1 == 0 & Q5_1 == 1, 'admin', 'none'))))
df %>% dplyr::select(Q0_DBN, supervisor_cert) %>% 
  group_by(supervisor_cert) %>% 
  count() %>% 
  ggplot(aes(x=supervisor_cert, y=n)) +
  geom_bar(stat='identity')
```

Few supervisors are certified in the arts, the majority are administrators. Could this have an impact on the efficacy of an arts program?

```{r supervisor cert stats, results = 'hide'}
df <- df %>%
  mutate(supervisor_cert_both = ifelse(Q4_1 == 1 & Q5_1 == 1, 1, 0))
q_certs <- c('Q4_1', 'Q5_1', 'supervisor_cert_both')
df %>%
  dplyr::select(q_certs, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
df %>%
  dplyr::select(q_certs, perc_34_all_2018_math) %>% 
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
```

Running a linear regression does not yield a statistically significant result for the effect of supervisor certification on student academic performance.

```{r supervisor cert viz}
df %>%
  dplyr::select(supervisor_cert, perc_34_all_2018_ela) %>% 
  ggplot(aes(x=perc_34_all_2018_ela, y=supervisor_cert)) +
  geom_density_ridges(scale = 1)
```

We can see greater density at the high end of academic performance for schools with supervisors who have both certifications, but there is no great difference in the shapes of the different groups.

### Resources devoted to arts education

Schools were asked how many rooms they have dedicated to arts education. This may be useful for assessing the resources that are made available for arts education more broadly.

```{r rooms setup}
q8 <- df %>% colnames() %>% .[grepl('Q8_', .)]
df <- df %>%
  mutate(rm_dance = Q8_R1_C1 + Q8_R1_C2,
         rm_music = Q8_R2_C1 + Q8_R2_C2,
         rm_thtr = Q8_R3_C1 + Q8_R3_C2,
         rm_media = Q8_R4_C1 + Q8_R4_C2)
rooms <- df %>% colnames %>% .[grepl('rm_', .)]
df %>% 
  dplyr::select(starts_with('rm')) %>% 
  gather(key = discipline,
         value = num) %>%
  ggplot(aes(x=num, color=discipline)) +
  geom_density()
#There is the greatest variation in music rooms
```

```{r rooms stats}
#ELA
df %>% dplyr::select(rooms, total_enrollment_2017, perc_34_all_2018_ela) %>% 
  lm(perc_34_all_2018_ela~., data=.) %>% 
  summary()
#Math
df %>% dplyr::select(rooms, total_enrollment_2017, perc_34_all_2018_math) %>%
  lm(perc_34_all_2018_math~., data=.) %>% 
  summary()
```

Conditioning on the effect of school size, through total enrollment, there is no statistically significant effect of the number of rooms for arts education on student academic performance.

We can also consider the technological resources available to students. Here we are referring to the various software, computers, cameras, and so on that can be made available for use in the arts. We do not have data on musical instruments for music or props for theater, although it is certainly the case that those resources are important in running an effective arts program.

```{r tools setup}
#Q9 technology tools available to students
q9 <- df %>% colnames() %>% .[grepl('Q9_', .)]
df %>% dplyr::select(q9) %>% colMeans() %>% 
  as_data_frame() %>% 
  ggplot(aes(x=reorder(q9, value), y=value)) +
  geom_bar(stat='identity')
#Smartboard is the most frequent tool
df %>% dplyr::select(q9, perc_34_4_2018_ela) %>% 
  lm(perc_34_4_2018_ela~., data=.) %>% 
  summary()
#_1 Animation software has a significant effect.
#_2 color printers, _4 digital tablets, _19 film cameras have a lesser but still significant effect.
#Does this effect remain if we control for wealth in some form, since these seem like expensive items?
df %>% dplyr::select(q9, perc_34_4_2018_math) %>% 
  lm(perc_34_4_2018_math~., data=.) %>% 
  summary()
#_4 digital tablets have a significant effect.
#_1 animation software has a lesser but significant effect.
```

```{r boroughs performance}
#Extract boroughs
df$boro <- df$Q0_DBN %>% str_extract('\\D')
df$boro %>% unique()
df <- df %>% mutate(
  K = if_else(boro == 'K', 1, 0),
  X = if_else(boro == 'X', 1, 0),
  M = if_else(boro == 'M', 1, 0),
  Q = if_else(boro == 'Q', 1, 0),
  R = if_else(boro == 'R', 1, 0)
  )
df %>% select(perc_34_4_2018_ela, K, X, M, Q, R) %>% 
  lm(perc_34_4_2018_ela~., data=.) %>% 
  summary()
#Bronx has a strong negative relation and Brooklyn a lesser negative relation. All other boroughs are not statistically significant.
df %>% select(perc_34_4_2018_math, K, X, M, Q, R) %>% 
  lm(perc_34_4_2018_math~., data=.) %>% 
  summary()
#Bronx has a strong negative correlation and no other borough has a statistically significant relation.
df %>% select(perc_34_4_2018_ela, perc_black_2017, perc_pov_2017, K, X, M, Q, R) %>% 
  lm(perc_34_4_2018_ela~., data=.) %>% 
  summary()
#Controlling for the percentage of black students and of poor students removed the statistical significance of the negative coefficient for the Bronx, and results in a positive statistically significant effect for Brooklyn.
```

```{r boroughs arts liaison}
df %>% select(Q2_1) %>% sum()
boros <- c('K', 'X', 'M', 'Q', 'R')
df %>% select(Q2_1, boros) %>% 
  lm(Q2_1~., data=.) %>% summary()
#Brooklyn and Manhattan schools have a statistically significant decreased chance of having an arts liaison.
df %>% select(perc_pov_2017, perc_black_2017, Q2_1, boros) %>% 
  lm(Q2_1~., data=.) %>% summary()
#Again, conditioning on poverty and share of black students removes the statistical significance.
```

```{r 4th grade hrs}
#Q16 arts instructional hours for fourth-grade
q16 <- df %>% colnames() %>% .[grepl('Q16_', .)]
df <- df %>% rowwise() %>% 
  mutate(
    dance_hr = sum(Q16_C1_R1 + Q16_C2_R1 ,  Q16_C3_R1 ,  Q16_C4_R1 ,  Q16_C5_R1 ,  Q16_C6_R1 ,  Q16_C7_R1 ,  Q16_C8_R1 ,  Q16_C9_R1 ,  Q16_C10_R1 ,  Q16_C11_R1 ,  Q16_C12_R1 ,  Q16_C13_R1 ,  Q16_C14_R1, na.rm = T),
    music_hr = sum(Q16_C1_R2 ,  Q16_C2_R2 ,  Q16_C3_R2 ,  Q16_C4_R2 ,  Q16_C5_R2 ,  Q16_C6_R2 ,  Q16_C7_R2 ,  Q16_C8_R2 ,  Q16_C9_R2 ,  Q16_C10_R2 ,  Q16_C11_R2 ,  Q16_C12_R2 ,  Q16_C13_R2 ,  Q16_C14_R2, na.rm = T),
    thtr_hr = sum(Q16_C1_R3 ,  Q16_C2_R3 ,  Q16_C3_R1 ,  Q16_C4_R1 ,  Q16_C5_R3 ,  Q16_C6_R3 ,  Q16_C7_R3 ,  Q16_C8_R3 ,  Q16_C9_R3 ,  Q16_C10_R3 ,  Q16_C11_R3 ,  Q16_C12_R3 ,  Q16_C13_R3 ,  Q16_C14_R1, na.rm = T),
    visart_hr = sum(Q16_C1_R4 ,  Q16_C2_R4 ,  Q16_C3_R4 ,  Q16_C4_R4 ,  Q16_C5_R4 ,  Q16_C6_R4 ,  Q16_C7_R4 ,  Q16_C8_R4 ,  Q16_C9_R4 ,  Q16_C10_R4 ,  Q16_C11_R4 ,  Q16_C12_R4 ,  Q16_C13_R4 ,  Q16_C14_R4, na.rm = T)
  )
q_4th_hrs <- df %>% colnames() %>% .[grepl('_hr', .)]
g1 <- df %>% dplyr::select(q_4th_hrs, perc_34_4_2018) %>% 
  ggplot(aes(x = dance_hr, y = perc_34_4_2018_ela)) +
  geom_point() + xlim(c(0, 600))
g2 <- df %>% dplyr::select(q_4th_hrs, perc_34_4_2018) %>% 
  ggplot(aes(x = music_hr, y = perc_34_4_2018_ela)) +
  geom_point() + xlim(c(0, 600))
g3 <- df %>% dplyr::select(q_4th_hrs, perc_34_4_2018) %>% 
  ggplot(aes(x = thtr_hr, y = perc_34_4_2018_ela)) +
  geom_point() + xlim(c(0, 600))
g4 <- df %>% dplyr::select(q_4th_hrs, perc_34_4_2018) %>% 
  ggplot(aes(x = visart_hr, y = perc_34_4_2018_ela)) +
  geom_point() + xlim(c(0, 600))
figure <- ggarrange(g1, g2, g3, g4,
                    labels = c('1', '2', '3', '4'),
                    ncol = 2,
                    nrow = 2)
figure
#No particular relation between hrs and perc_34
df %>% dplyr::select(q_4th_hrs, perc_34_4_2018_ela) %>% 
  lm(perc_34_4_2018_ela~., data=.) %>% 
  summary()
df %>% dplyr::select(q_4th_hrs, perc_34_4_2018_math) %>% 
  lm(perc_34_4_2018_math~., data=.) %>% 
  summary()
#Visual arts has a barely significant relation with performance.
```

```{r screened arts}
df %>% dplyr::select(Q20_1) %>% sum()
```

Only 27 screened arts programs in NYC schools.

```{r 3yr sequence}
q22 <- df %>% colnames() %>% .[grepl('Q22', .)] %>% .[grepl('_C1', .)]
df %>% dplyr::select(q22) %>% colSums() %>% 
  as_data_frame() %>% 
  ggplot(aes(x=disc_5, y=value)) +
  geom_bar(stat='identity')
#Many visual arts and music sequences. Fewer dance and theater, and very few film.
df %>% dplyr::select(q22, perc_34_7_2018_ela) %>% 
  lm(perc_34_7_2018_ela~., data=.) %>% 
  summary()
df %>% dplyr::select(q22, perc_34_7_2018_math) %>% 
  lm(perc_34_7_2018_math~., data=.) %>% 
  summary()
#Music programs have a positive effect on academic performance in 7th grade.
```

### Funding

```{r arts funding sources}
q36 <- q_arts %>% .[grepl('Q36', .)] %>% .[grepl('C1', .)]
fund_srcs <- c('Cultural organizations',
               'Education association',
               'Federal, state, or city grants',
               'Local business or corporation',
               'Private foundation',
               'PTA/PA',
               'State, county, local arts councils')
df %>% dplyr::select(q36) %>% colSums() %>% 
  as_data_frame() %>% 
  ggplot(aes(x=reorder(fund_srcs, -value), y=value)) +
  geom_bar(stat='identity') +
  coord_flip()
```

Government grants are the most common source of non-DOE funding, followed by cultural organizations, and PTA. Then arts councils, private foundations, educational associations and local business.

Does poverty predict the different sources of funding?
```{r art funding poverty}
for (i in q36) {
  temp <- df %>% dplyr::select(i, perc_pov_2017)
  temp.glm <- glm(unlist(temp[,1])~unlist(temp[,2]), family='binomial')
  print(i); print(summary(temp.glm))
}
```

Sources that are not predicted with statistical significance: Cultural organizations, education associations, government grants, local business, private foundations.  
Poverty is negatively associated with PTA/PA at a *** sig. Local arts councils is similar but at a * sig. This could be picking up the effect of specific neighborhoods?

```{r art funding poverty ctrl black}
for (i in q36) {
  temp <- df %>% dplyr::select(i, perc_pov_2017, perc_black_2017)
  temp.glm <- glm(unlist(temp[,1])~unlist(temp[,2])+unlist(temp[,3]), family='binomial')
  print(i); print(summary(temp.glm))
}
```

Conditioning on poverty and percentage of black students yields a positive relationship with poverty for government grants and a negative relationship with poverty for PTA/PA funding (negative with black students for both).

```{r art funding poverty ctrl other}
for (i in q36) {
  temp <- df %>% dplyr::select(i, perc_pov_2017, perc_black_2017, total_enrollment_2017, perc_34_all_2018)
  temp.glm <- glm(unlist(temp[,1])~unlist(temp[,2])+unlist(temp[,3])+unlist(temp[,4])+unlist(temp[,5])+unlist(temp[,6]), family='binomial')
  print(i); print(summary(temp.glm))
}
```

Controlling for poverty, percentage of black students, total enrollment, and academic performance yields a statistically significant negative relationship for poverty and a positive for math performance on PTA/PA funding, and a positive relationship for total enrollment on state funding.

### Parental Involvement
```{r parental inv setup}
q39 <- q_arts %>% .[grepl('Q39', .)]
par_inv <- c('Attending school arts events',
             'Volunteering in arts programs',
             'Donating arts materials',
             'Other')
df %>% dplyr::select(Q39_4, Q0_DBN) %>% 
  group_by(Q39_4) %>% count()
```

It could be neat to analyze the 'Other' category. For now I'll move the comments over and replace them with 1s.

```{r parental inv mutate}
df$Q39_Other <- df$Q39_4
df <- df %>% 
  mutate(Q39_4 = ifelse(Q39_4 == 0, 0, 1))
#Move 'Other' over to another column
df %>% dplyr::select(q39) %>% colSums() %>% 
  as_data_frame() %>% 
  ggplot(aes(x=reorder(par_inv, -value), y=value)) +
  geom_bar(stat='identity') +
  coord_flip()
```

Parental participation is most common through attending school events, to be expected. Fewer parents volunteer or donate art materials.  

Is there a relationship between poverty and volunteering in arts programs, Q39_2?

```{r parental inv volunteer}
df %>% 
  dplyr::select(Q39_2, perc_pov_2017, total_enrollment_2017, perc_34_all_2018) %>% 
  glm(Q39_2~., data=., family='binomial') %>% 
  summary()
```

There is a strong negative relation between poverty and parental volunteers in arts programs, even controlling for school size and academic performance.  

Is there a relationship between poverty and donating arts materials, Q39_3?

```{r parental inv donation}
df %>% select(Q39_3, perc_pov_2017, total_enrollment_2017, perc_34_all_2018) %>% 
  glm(Q39_3~., data=., family='binomial') %>% 
  summary()
```

There is a strong negative relation between poverty and donating arts materials and a postiive relation between math scores and donations. Odd.

### Arts opportunities at the school and outside of school

```{r arts opportunities setup}
q40 <- q_arts %>% .[grepl('Q40', .)]
arts_opp <- c('Artwork exhibits',
              'Concerts',
              'Dance performances',
              'Films',
              'Theater performances',
              'Other')
arts_opp <- factor(arts_opp,
                   levels=c('Artwork exhibits',
              'Concerts',
              'Dance performances',
              'Films',
              'Theater performances',
              'Other'))
q40_school <- q40 %>% .[grepl('C1', .)]
q40_outside <- q40 %>% .[grepl('C2', .)]
df %>% dplyr::select(q40_school) %>% colSums(na.rm = T) %>% 
  as_data_frame() %>% 
  ggplot(aes(x=reorder(arts_opp, -value), y=value)) +
  geom_bar(stat='identity') + coord_flip()
```

At school, concerts, dance, theater performances and art exhibits are of decreasing frequency. Films are shown less frequently.

```{r arts opportunities by location}
g1 <- df %>% select(q40_outside) %>% colSums(na.rm = T) %>% 
  as_data_frame() %>% 
  ggplot(aes(x=arts_opp, y=value)) +
  geom_bar(stat='identity') + coord_flip()
g2 <- df %>% select(q40_school) %>% colSums(na.rm = T) %>% 
  as_data_frame() %>% 
  ggplot(aes(x=arts_opp, y=value)) +
  geom_bar(stat='identity') + coord_flip()
figure <- ggarrange(g1, g2,
                    labels = c('Out', 'In'),
                    ncol = 2)
figure
```

Dance performances and concerts tend to be in schools, theater performances are more likely to be outside of school.

Which is a better predictor of school events, wealth or having an arts program?



To do matching, I should have a better idea of what I am considering "treatment", "control" and "outcome".
```{r tools matchit}
f_ela <- as.formula(paste('good_all_2018_ela~', paste(q9, collapse='+')))
f_math <- as.formula(paste('good_all_2018_math~', paste(q9, collapse='+')))
temp <- df %>%
  dplyr::select(q9, good_all_2018_ela) %>%
  filter(!is.na(good_all_2018_ela))
```

### Data Provenance
The NYC Department of Education (DOE) has released an annual arts survey report for several years. After separate csv files were downloaded for each source, I changed performance and demographic data into wide format and merged them with the arts survey data on school District Borough Number (DBN).  